---
title: "Introduction to bionmf"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to bionmf}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Overview

**bionmf** fits non-negative matrix factorization (NMF) to a
gene-by-sample expression matrix. The model is

$$
X \approx W H, \qquad W \ge 0,\ H \ge 0.
$$

-   Rows of $X$ are genes; columns are samples.
-   Columns of $W$ are latent **gene programs**.
-   Rows of $H$ are **sample loadings** on those programs.

The fit uses Lee–Seung multiplicative updates under the Frobenius loss,
implemented in base R.

```{r setup}
library(bionmf)
```

# Load the bundled example

The package ships a simulated matrix (50 genes $\times$ 100 samples,
with 3 distinct expression patterns) as
`inst/extdata/expression_matrix.csv`. Use `expression_matrix_path()` so
the file is found after install, without hard-coded working-directory
paths.

```{r load}
path <- expression_matrix_path()
X <- load_expression_matrix(path)
dim(X)
```

`load_expression_matrix()` checks that values are non-missing and
non-negative, which NMF requires.

# Fit NMF

Choose a factorization rank (number of programs), a maximum number of
iterations, and a relative-error tolerance. A seed makes the random
initialization reproducible.

```{r fit}
fit <- run_nmf(
  X,
  rank = 3L,
  max_iter = 200L,
  tol = 1e-4,
  seed = 42L
)

fit$iterations
```

The result is a list with factor matrices and the error trajectory:

-   `fit$W` — genes $\times$ factors
-   `fit$H` — factors $\times$ samples
-   `fit$error_history` — Frobenius error after each iteration
-   `fit$iterations` — number of updates performed (may stop early)

```{r dims}
dim(fit$W)
dim(fit$H)
```

# Diagnostics

Reconstruction quality is summarized by the Frobenius residual
$\|X - WH\|_F$ and by the fraction of variance explained.

```{r diagnostics}
reconstruction_error(X, fit$W, fit$H)
explain_variance(X, fit$W, fit$H)
```

# Plots

The convergence curve should decrease and then flatten when the relative
change in error falls below `tol`.

```{r error-plot}
plot_error_history(fit$error_history)
```

Heatmaps of $W$ and $H$ show which genes load on each program and how
samples use those programs.

```{r heatmaps, fig.width = 10}
plot_nmf_heatmaps(fit$W, fit$H)
```

# Using your own data

Pass any gene-by-sample CSV with gene IDs in the first column:

```{r custom, eval = FALSE}
## NOTE: comment out the below code and replace with your dataset path

 
# X <- load_expression_matrix("path/to/your_expression.csv")
# fit <- run_nmf(X, rank = 4L, seed = 1L)
```

See `?run_nmf`, `?load_expression_matrix`, and `?explain_variance` for
argument details.
